---
title: "Lab1"
author: "Bingqing Li"
format: pdf
editor: visual
---

```{r}
#| message: false

library(tidyverse)
library(ggplot2)
```

```{r}
#| warning: FALSE
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", 
                 skip = 2, col_types = "dcddd")
head(dm)
```

## Q1
1. Plot the ratio of female to male mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme 
```{r}
dm |>
  filter(Age %in% c(10, 20, 30, 40)) |>
  mutate(fm_ratio = Female/Male) |> 
  ggplot(aes(x = Year, y = fm_ratio, color = Age)) +
  geom_line() +
  theme_bw() +
  labs(title = 'The ratio of female to male mortality rates') +
  theme(plot.title = element_text(hjust = 0.5))


```

## Q2
2. Find the age that has the lowest female mortality rate each year 
```{r}
dm |>
  group_by(Year) |>
  summarise(lf_age = Age[which.min(Female)])
```

## Q3
3. Use the `summarize(across())` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations. 
```{r}
#| warning: FALSE
dm |>
  group_by(Age) |>
  summarise(across(Female:Total, sd, na.rm = TRUE)) |>
  arrange(as.numeric(Age))
```

## Q4
4. The Canadian HMD also provides population sizes over time (https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt). Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: `left_join` will probably be useful here. 

```{r}
#| warning: FALSE
dp <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt",
                 skip = 2, col_types = "dcddd")
head(dp)
```

```{r}
#| warning: false
dl <- left_join(dm, dp, by=c('Year', 'Age'))
colnames(dl) <- c('Year','Age', 'FemaleM', 'MaleM', 
                  'TotalM', 'FemaleP', 'MaleP', 'TotalP' )
head(dl)
dl1 <- dl |>
  group_by(Year) |>
  summarise(weighted_mean_F = sum(FemaleM*FemaleP, 
                                  na.rm = TRUE) /sum(FemaleP, na.rm = TRUE),
            weighted_mean_M =  sum(MaleM*MaleP, 
                                   na.rm = TRUE) /sum(MaleP, na.rm = TRUE) )|>
  pivot_longer(weighted_mean_F:weighted_mean_M, names_to = 'Sex', 
               values_to = 'Weighted_Mean') |>
  mutate(Sex = case_when(
    Sex == 'weighted_mean_F' ~ 'Female',
    Sex == 'weighted_mean_M' ~ 'Male',
    TRUE ~ as.character(Sex)
  )) 

dl1
```

```{r}
dl1 |>
  ggplot(aes(x = Year, y = Weighted_Mean, color = Sex)) + 
  geom_line() +
  theme_bw() + 
  labs(y = 'Weighted Mortality',
       title = 'The population weighted average mortality rate') +
  theme(plot.title = element_text(hjust = 0.5))
```
The plot indicates that over the years, females have exhibited a lower average mortality rate compared to males, and   the mortality rate for male and female shows a general downward trend.

## Q5
5. Write down using appropriate notation, and run a simple linear regression with logged mortality rates as the outcome and age (as a continuous variable) as the covariate, using data for females aged less than 106 for the year 2000. Interpret the coefficient on age. 

```{r}
dm$Age <- as.numeric(dm$Age)
dm1 <- dm |>
  filter(Year == 2000 & Age < 106 & !is.na(Female)) |>
  select(Year, Age, Female)
dm1

```

```{r}
model <-  lm(log(Female)~ Age, data = dm1)
coef(model)


p <- ggplot(dm1, aes(x = Age, y = log(Female))) +
  geom_point() + 
  geom_smooth(method = "lm", color = "red") +  
  labs(
    x = "Age",
    y = "Log of Mortality Rate",
    title = "Scatter Plot with Fitted Regression Line"
  ) +
  theme(plot.title = element_text(hjust = 0.5))+
  theme_bw()

p
```

Because $$log(Female's \ Mortality_i) = 0.087\times Age_i -10.062,$$
then $$Female's \ Mortality_i = e^{0.087\times Age_i -10.062}.$$ Thus, for females aged less than 106 for the year 2000, the mortality rate increases with age. Expected value change $e^{0.087}$ in Mortality rate with one unit increase in Age.

